---
title: "Research Analytics"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 7, dpi = 300)

# Load required libraries
library(dplyr)
library(ggplot2)
library(DT)
library(stringr)
library(lubridate)
library(tidyr)

# Load the statistics
if(file.exists("../website_stats.rds")) {
  stats <- readRDS("../website_stats.rds")
} else {
  source("../R/generate_website_stats.R")
  stats <- readRDS("../website_stats.rds")
}

all_data <- stats$pub_data
```

# Research Analytics Dashboard

Comprehensive analysis of research trends, collaboration patterns, and impact metrics for the School of Earth and Sustainability.

---

## Research Output Trends

```{r combined-trends}
# Combined faculty and student trends
yearly_combined <- all_data %>%
  filter(Year >= 2009) %>%
  group_by(Year) %>%
  summarise(
    Total_Publications = n(),
    Faculty_Only = sum(SES_Faculty != "" & SES_Grad_Students == "", na.rm = TRUE),
    With_Students = sum(SES_Faculty != "" & SES_Grad_Students != "", na.rm = TRUE),
    Student_Only = sum(SES_Faculty == "" & SES_Grad_Students != "", na.rm = TRUE),
    Total_Citations = sum(Citations, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  pivot_longer(
    cols = c(Faculty_Only, With_Students, Student_Only),
    names_to = "Publication_Type",
    values_to = "Count"
  ) %>%
  mutate(
    Publication_Type = case_when(
      Publication_Type == "Faculty_Only" ~ "Faculty Only",
      Publication_Type == "With_Students" ~ "Faculty + Students",
      Publication_Type == "Student_Only" ~ "Students Only"
    )
  )

p1 <- ggplot(yearly_combined, aes(x = Year, y = Count, fill = Publication_Type)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c")) +
  labs(
    title = "Publication Trends by Author Type",
    subtitle = "Stacked area chart showing collaboration patterns over time",
    x = "Year",
    y = "Number of Publications",
    fill = "Author Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )

print(p1)
```

---

## Research Impact Over Time

```{r impact-analysis}
# Calculate cumulative metrics
impact_data <- all_data %>%
  filter(Year >= 2009, SES_Faculty != "") %>%
  arrange(Year) %>%
  mutate(
    Cumulative_Pubs = cumsum(!duplicated(paste(Title, Authors))),
    Cumulative_Citations = cumsum(Citations)
  ) %>%
  group_by(Year) %>%
  summarise(
    Annual_Pubs = n(),
    Annual_Citations = sum(Citations),
    Cumulative_Pubs = max(Cumulative_Pubs),
    Cumulative_Citations = max(Cumulative_Citations),
    .groups = 'drop'
  )

# Create static ggplot
p2 <- ggplot(impact_data, aes(x = Year)) +
  geom_col(aes(y = Annual_Pubs), fill = "#1f77b4", alpha = 0.8) +
  labs(
    title = "Research Output vs. Cumulative Impact",
    subtitle = "Annual publications with cumulative citations trend",
    x = "Year",
    y = "Annual Publications"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold")
  )

print(p2)
```

---

## Faculty Collaboration Network

```{r collaboration-analysis}
# Analyze multi-faculty collaborations
collab_pubs <- all_data %>%
  filter(SES_Faculty != "", str_count(SES_Faculty, ",") >= 1) %>%  # Multi-faculty papers
  mutate(faculty_list = strsplit(SES_Faculty, ", ")) %>%
  select(Title, Year, Citations, faculty_list)

# Count collaborations
faculty_collabs <- collab_pubs %>%
  mutate(num_faculty = lengths(faculty_list)) %>%
  group_by(num_faculty) %>%
  summarise(
    Publications = n(),
    Avg_Citations = round(mean(Citations, na.rm = TRUE), 1),
    Total_Citations = sum(Citations, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(Faculty_Group = paste(num_faculty, "Faculty"))

p3 <- ggplot(faculty_collabs, aes(x = reorder(Faculty_Group, num_faculty), y = Publications)) +
  geom_col(fill = "#9467bd", alpha = 0.8) +
  geom_text(aes(label = Publications), vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Multi-Faculty Collaborative Publications",
    subtitle = "Publications with multiple SES faculty co-authors",
    x = "Number of SES Faculty Co-Authors",
    y = "Number of Publications"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 11)
  )

print(p3)
```

---

## Citation Distribution Analysis

```{r citation-distribution}
# Citation distribution
citation_bins <- all_data %>%
  filter(SES_Faculty != "", Citations > 0) %>%
  mutate(
    Citation_Range = case_when(
      Citations <= 5 ~ "1-5",
      Citations <= 10 ~ "6-10", 
      Citations <= 25 ~ "11-25",
      Citations <= 50 ~ "26-50",
      Citations <= 100 ~ "51-100",
      Citations <= 200 ~ "101-200",
      TRUE ~ "200+"
    )
  ) %>%
  count(Citation_Range) %>%
  mutate(
    Citation_Range = factor(Citation_Range, levels = c("1-5", "6-10", "11-25", "26-50", "51-100", "101-200", "200+"))
  )

p4 <- ggplot(citation_bins, aes(x = Citation_Range, y = n)) +
  geom_col(fill = "#2ca02c", alpha = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Distribution of Publication Citations",
    subtitle = "Number of publications in each citation range",
    x = "Citation Range",
    y = "Number of Publications"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 11)
  )

print(p4)
```

---

## Student Mentorship Impact

```{r student-mentorship}
# Analyze publications with both faculty and students
mentorship_data <- all_data %>%
  filter(SES_Faculty != "", SES_Grad_Students != "") %>%
  group_by(Year) %>%
  summarise(
    Mentorship_Pubs = n(),
    Avg_Citations = round(mean(Citations, na.rm = TRUE), 1),
    Total_Students = length(unique(unlist(strsplit(paste(SES_Grad_Students, collapse = ", "), ", ")))),
    .groups = 'drop'
  )

p5 <- ggplot(mentorship_data, aes(x = Year)) +
  geom_col(aes(y = Mentorship_Pubs), fill = "#ff7f0e", alpha = 0.8) +
  labs(
    title = "Faculty-Student Collaborative Publications",
    subtitle = "Publications with student co-authors by year",
    x = "Year",
    y = "Publications with Students"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold")
  )

print(p5)
```

---

## Top Research Keywords

```{r keyword-analysis}
# Extract keywords from titles (simple approach)
# This is a basic implementation - could be enhanced with proper NLP
common_words <- c("the", "a", "an", "and", "or", "but", "in", "on", "at", "to", "for", "of", "with", "by", "from", "as", "is", "are", "was", "were", "been", "be", "have", "has", "had", "do", "does", "did", "will", "would", "could", "should", "may", "might", "can", "must")

keywords <- all_data %>%
  filter(SES_Faculty != "", !is.na(Title)) %>%
  mutate(
    # Extract potential keywords from titles
    title_words = str_split(tolower(Title), "[^a-zA-Z]+")
  ) %>%
  select(Year, title_words) %>%
  tidyr::unnest(title_words) %>%
  filter(
    !title_words %in% common_words,
    nchar(title_words) > 3,
    !str_detect(title_words, "^[0-9]+$")
  ) %>%
  count(title_words, sort = TRUE) %>%
  top_n(20, n) %>%
  mutate(title_words = str_to_title(title_words))

p6 <- ggplot(keywords, aes(x = reorder(title_words, n), y = n)) +
  geom_col(fill = "#17becf", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Most Common Words in Publication Titles",
    subtitle = "Top 20 research-related terms (simplified analysis)",
    x = "Term",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(p6)
```

---

## Department Performance Metrics

```{r performance-dashboard}
# Calculate key performance indicators
current_year <- year(Sys.Date())
recent_years <- (current_year-5):current_year

kpi_data <- all_data %>%
  filter(SES_Faculty != "") %>%
  summarise(
    Total_Pubs = n(),
    Recent_Pubs = sum(Year %in% recent_years, na.rm = TRUE),
    Total_Citations = sum(Citations, na.rm = TRUE),
    Avg_Citations_Per_Pub = round(mean(Citations, na.rm = TRUE), 1),
    High_Impact = sum(Citations >= 50, na.rm = TRUE),
    Collaboration_Rate = round(100 * mean(str_count(SES_Faculty, ",") > 0, na.rm = TRUE), 1),
    Student_Mentorship_Rate = round(100 * mean(SES_Grad_Students != "", na.rm = TRUE), 1),
    Active_Faculty = stats$actively_publishing_count
  )

# Create a summary table
kpi_table <- data.frame(
  Metric = c(
    "Total Publications (2009+)",
    "Publications (Last 5 Years)", 
    "Total Citations",
    "Average Citations per Publication",
    "High-Impact Publications (50+ citations)",
    "Multi-Faculty Collaboration Rate (%)",
    "Student Co-authorship  Rate (%)",
    "Actively Publishing Faculty (since 2023)",
    "Department H-Index"
  ),
  Value = c(
    format(kpi_data$Total_Pubs, big.mark = ","),
    format(kpi_data$Recent_Pubs, big.mark = ","),
    format(kpi_data$Total_Citations, big.mark = ","),
    kpi_data$Avg_Citations_Per_Pub,
    format(kpi_data$High_Impact, big.mark = ","),
    paste0(kpi_data$Collaboration_Rate, "%"),
    paste0(kpi_data$Student_Mentorship_Rate, "%"),
    kpi_data$Active_Faculty,
    stats$h_index
  )
)

knitr::kable(
  kpi_table,
  format = "html",
  caption = "Key Performance Indicators - SES Research Output",
  col.names = c("Metric", "Value"),
  align = c("l", "r"),
  escape = FALSE,
  table.attr = 'class="table table-striped table-hover" style="font-size: 13px; margin-left: auto; margin-right: auto;"'
)
```

---

## Annual Research Highlights

```{r yearly-highlights}
# Generate highlights for each recent year
highlights <- all_data %>%
  filter(SES_Faculty != "", Year >= (current_year - 4), Year <= current_year) %>%
  group_by(Year) %>%
  summarise(
    Publications = n(),
    Total_Citations = sum(Citations, na.rm = TRUE),
    Top_Citation = max(Citations, na.rm = TRUE),
    Student_Collaborations = sum(SES_Grad_Students != "", na.rm = TRUE),
    Faculty_Count = length(unique(unlist(strsplit(paste(SES_Faculty, collapse = ", "), ", ")))),
    Top_Journal = names(sort(table(Journal), decreasing = TRUE))[1],
    .groups = 'drop'
  ) %>%
  arrange(desc(Year))

knitr::kable(
  highlights,
  format = "html", 
  caption = "Annual Research Highlights (Last 5 Years)",
  col.names = c('Year', 'Publications', 'Total Citations', 'Highest Citation', 'Student Collaborations', 'Active Faculty', 'Top Journal'),
  align = c("c", "r", "r", "r", "r", "r", "l"),
  escape = FALSE,
  table.attr = 'class="table table-striped table-hover" style="font-size: 12px; margin-left: auto; margin-right: auto;"'
)
```

---

*Analytics generated from `r format(nrow(all_data), big.mark = ",")` publications spanning `r min(all_data$Year)` to `r max(all_data$Year)`. Last updated: `r format(Sys.time(), "%B %d, %Y")`*