---
title: "Student Publications"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 7, dpi = 300)

# Load required libraries
library(dplyr)
library(ggplot2)
library(DT)
library(stringr)
library(lubridate)
library(tidyr)

# Load the statistics
if(file.exists("../website_stats.rds")) {
  stats <- readRDS("../website_stats.rds")
} else {
  source("../R/generate_website_stats.R")
  stats <- readRDS("../website_stats.rds")
}

student_data <- stats$grad_pubs
```

# Student Publications Dashboard

Since 2009, **`r stats$total_grad_students`** SES graduate students have published **`r format(nrow(student_data), big.mark = ",")`** peer-reviewed articles as co-authors, accumulating **`r format(sum(student_data$Citations, na.rm = TRUE), big.mark = ",")`** total citations.

*The "Most Cited Publications" table below highlights research where graduate students served as **first authors**, showcasing their leadership in conducting and publishing original research.*

---

## Graduate Student Publishing Trends

```{r student-pubs-by-year}
# Publications by year with student counts
yearly_student_pubs <- student_data %>%
  filter(Year >= 2009) %>%
  group_by(Year) %>%
  summarise(
    Publications = n(),
    Citations = sum(Citations, na.rm = TRUE),
    # Count unique students per year
    Student_Authors = length(unique(unlist(strsplit(paste(SES_Grad_Students, collapse = ", "), ", ")))),
    .groups = 'drop'
  )

p1 <- ggplot(yearly_student_pubs, aes(x = Year)) +
  geom_col(aes(y = Publications), fill = "#1f77b4", alpha = 0.8) +
  geom_line(aes(y = Student_Authors), color = "#ff7f0e", size = 1.5) +
  geom_point(aes(y = Student_Authors), color = "#ff7f0e", size = 3) +
  labs(
    title = "Graduate Student Publications by Year",
    subtitle = "Blue bars: Publications | Orange line: Number of students publishing",
    x = "Year",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold")
  )

print(p1)
```

---

## Student Publication Impact

```{r student-citations}
p2 <- ggplot(yearly_student_pubs, aes(x = Year, y = Citations)) +
  geom_col(fill = "#2ca02c", alpha = 0.8) +
  labs(
    title = "Graduate Student Publication Citations by Year",
    subtitle = "Total citations for publications with graduate student authors",
    x = "Year",
    y = "Total Citations"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold")
  )

print(p2)
```

---

<!-- ## Most Productive Graduate Students -->

<!-- ```{r student-productivity} -->
<!-- # Count publications per student -->
<!-- student_productivity <- student_data %>% -->
<!--   filter(SES_Grad_Students != "") %>% -->
<!--   # Split student names and count -->
<!--   mutate(student_list = strsplit(SES_Grad_Students, ", ")) %>% -->
<!--   unnest(student_list) %>% -->
<!--   group_by(student_list) %>% -->
<!--   summarise( -->
<!--     Publications = n(), -->
<!--     Total_Citations = sum(Citations, na.rm = TRUE), -->
<!--     Avg_Citations = round(mean(Citations, na.rm = TRUE), 1), -->
<!--     Years_Active = paste(sort(unique(Year)), collapse = ", "), -->
<!--     .groups = 'drop' -->
<!--   ) %>% -->
<!--   arrange(desc(Publications)) %>% -->
<!--   top_n(20, Publications)  # Top 20 most productive -->

<!-- p3 <- ggplot(student_productivity, aes(x = reorder(student_list, Publications), y = Publications)) + -->
<!--   geom_col(fill = "#9467bd", alpha = 0.8) + -->
<!--   coord_flip() + -->
<!--   labs( -->
<!--     title = "Top 20 Most Productive Graduate Students (2009+)", -->
<!--     subtitle = "Number of publications per student", -->
<!--     x = "Graduate Student", -->
<!--     y = "Number of Publications" -->
<!--   ) + -->
<!--   theme_minimal() + -->
<!--   theme( -->
<!--     plot.title = element_text(size = 16, face = "bold"), -->
<!--     plot.subtitle = element_text(size = 12), -->
<!--     axis.text.y = element_text(size = 9), -->
<!--     axis.text.x = element_text(size = 10), -->
<!--     axis.title = element_text(size = 12, face = "bold") -->
<!--   ) -->

<!-- ggplotly(p3, tooltip = c("x", "y")) %>% -->
<!--   layout( -->
<!--     xaxis = list(fixedrange = FALSE), -->
<!--     yaxis = list(fixedrange = FALSE) -->
<!--   ) %>% -->
<!--   config(displayModeBar = TRUE, displaylogo = FALSE) -->
<!-- ``` -->

<!-- --- -->

## Most Cited Publications with Graduate Students as First Authors

```{r top-cited-student-pubs}
# Function to extract first author's last name
get_first_author_lastname <- function(authors_string) {
  if(is.na(authors_string) || authors_string == "") return("")
  
  # Split by comma to get first author
  first_author <- trimws(strsplit(authors_string, ",")[[1]][1])
  
  # Handle different formats:
  # "J Smith" -> "Smith"  
  # "JA Smith" -> "Smith"
  # "Smith J" -> "Smith"
  # "Smith JA" -> "Smith"
  
  words <- trimws(unlist(strsplit(first_author, " ")))
  
  if(length(words) == 1) {
    # Single word, assume it's the last name
    last_name <- words[1]
  } else if(length(words) == 2) {
    # Two words - check which one looks like initials vs full name
    word1 <- words[1]
    word2 <- words[2]
    
    # If first word is short (1-3 chars) and second is longer, assume "J Smith" format
    if(nchar(word1) <= 3 && nchar(word2) > 3) {
      last_name <- word2
    } else {
      # Otherwise assume "Smith J" format
      last_name <- word1
    }
  } else {
    # More than 2 words, take the last one as surname
    last_name <- words[length(words)]
  }
  
  # Remove periods and clean up
  last_name <- gsub("[.]", "", last_name)
  
  return(last_name)
}

# Function to extract first author's first initial
get_first_author_initial <- function(authors_string) {
  if(is.na(authors_string) || authors_string == "") return("")
  
  # Split by comma to get first author
  first_author <- trimws(strsplit(authors_string, ",")[[1]][1])
  
  # Handle different formats to extract first initial:
  # "J Smith" -> "J"
  # "JA Smith" -> "J"  
  # "Smith J" -> "J"
  # "Smith JA" -> "J"
  
  words <- trimws(unlist(strsplit(first_author, " ")))
  
  if(length(words) == 1) {
    # Single word - assume it's last name only, no initial available
    return("")
  } else if(length(words) == 2) {
    word1 <- words[1]
    word2 <- words[2]
    
    # If first word is short (1-3 chars) and second is longer, assume "J Smith" format
    if(nchar(word1) <= 3 && nchar(word2) > 3) {
      initial <- substr(word1, 1, 1)  # Take first character of first word
    } else {
      # Otherwise assume "Smith J" format - take first character of second word
      initial <- substr(word2, 1, 1)
    }
  } else {
    # More than 2 words, try to find the shortest word as initials
    # Usually initials are shorter than names
    word_lengths <- nchar(words)
    shortest_idx <- which.min(word_lengths)
    initial <- substr(words[shortest_idx], 1, 1)
  }
  
  # Clean up and return
  initial <- gsub("[.]", "", initial)
  return(toupper(initial))
}

# Function to check if student is first author
is_student_first_author <- function(authors_string, student_names_string) {
  if(is.na(authors_string) || is.na(student_names_string) || 
     authors_string == "" || student_names_string == "") return(FALSE)
  
  first_author_lastname <- get_first_author_lastname(authors_string)
  first_author_initial <- get_first_author_initial(authors_string)
  
  # Split student names and check each one
  student_names <- trimws(unlist(strsplit(student_names_string, ",")))
  
  # For each student, extract their names and compare
  for(student in student_names) {
    if(student == "") next
    
    # Extract first name and last name from student
    if(grepl(",", student)) {
      # Format: "Last, First"
      parts <- trimws(unlist(strsplit(student, ",")))
      student_lastname <- parts[1]
      student_firstname <- if(length(parts) > 1) parts[2] else ""
    } else {
      # Format: "First Last" 
      words <- trimws(unlist(strsplit(student, " ")))
      student_firstname <- words[1]
      student_lastname <- words[length(words)]
    }
    
    # Remove periods and clean up
    student_lastname <- gsub("[.]", "", student_lastname)
    student_firstname <- gsub("[.]", "", student_firstname)
    
    # Get first initial of student's first name
    student_initial <- if(nchar(student_firstname) > 0) toupper(substr(student_firstname, 1, 1)) else ""
    
    # Check if both last name AND first initial match
    lastname_match <- tolower(first_author_lastname) == tolower(student_lastname)
    initial_match <- (first_author_initial == "" || student_initial == "" || first_author_initial == student_initial)
    
    if(lastname_match && initial_match) {
      return(TRUE)
    }
  }
  
  return(FALSE)
}

# Filter for publications where a graduate student is the first author
top_cited_students <- student_data %>%
  filter(SES_Grad_Students != "", !is.na(Citations), Citations > 0) %>%
  # Add a column to check if student is first author
  rowwise() %>%
  mutate(student_is_first_author = is_student_first_author(Authors, SES_Grad_Students)) %>%
  ungroup() %>%
  # Filter only publications where student is first author
  filter(student_is_first_author == TRUE) %>%
  arrange(desc(Citations)) %>%
  select(Title, Authors, Journal, Year, Citations, SES_Grad_Students) %>%
  head(20)

# Found data - proceeding with table

if(nrow(top_cited_students) > 0) {
  # Clean up long titles and author lists and rename column
  top_cited_students_clean <- top_cited_students %>%
    mutate(
      Title = ifelse(is.na(Title), "Unknown Title", Title),
      Authors = ifelse(is.na(Authors), "Unknown Authors", Authors),
      Journal = ifelse(is.na(Journal), "Unknown Journal", Journal),
      Title = ifelse(nchar(Title) > 80, paste0(substr(Title, 1, 77), "..."), Title),
      Authors = ifelse(nchar(Authors) > 100, paste0(substr(Authors, 1, 97), "..."), Authors),
      Journal = ifelse(nchar(Journal) > 40, paste0(substr(Journal, 1, 37), "..."), Journal)
    ) %>%
    rename(Student = SES_Grad_Students)  # Rename column header as requested
  
  # Create a simple HTML table that will definitely work
  top_cited_students_clean %>%
    knitr::kable(
      format = "html",
      caption = "Top 20 Most Cited Publications with Graduate Students as First Authors",
      col.names = c("Title", "Authors", "Journal", "Year", "Citations", "First Author Student"),
      align = c("l", "l", "l", "c", "r", "l"),
      escape = FALSE,
      table.attr = 'class="table table-striped table-hover" style="font-size: 12px; margin-left: auto; margin-right: auto;"'
    )
} else {
  cat("No citation data available for student publications.")
}
```

---

## Student Publication Journals

```{r student-journals}
# Top journals for student publications
student_journal_counts <- student_data %>%
  filter(!is.na(Journal), Journal != "") %>%
  count(Journal, sort = TRUE) %>%
  top_n(15, n)

p4 <- ggplot(student_journal_counts, aes(x = reorder(Journal, n), y = n)) +
  geom_col(fill = "#17becf", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Top 15 Journals for Graduate Student Publications",
    subtitle = "Number of publications per journal (with grad student co-authors)",
    x = "Journal",
    y = "Number of Publications"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )

print(p4)
```

---

## Summary Statistics

```{r student-summary-stats}
# Create student summary statistics table
student_summary <- data.frame(
  Metric = c(
    "Student Publications",
    "Total Citations",
    "Publishing Students", 
    "Avg Citations/Paper"
  ),
  Value = c(
    format(nrow(student_data), big.mark = ","),
    format(sum(student_data$Citations, na.rm = TRUE), big.mark = ","),
    stats$total_grad_students,
    round(mean(student_data$Citations, na.rm = TRUE), 1)
  )
)

knitr::kable(
  student_summary,
  format = "html",
  caption = "Student Research Summary Statistics",
  col.names = c("Metric", "Value"),
  align = c("l", "r"),
  escape = FALSE,
  table.attr = 'class="table table-striped table-hover" style="font-size: 14px; margin-left: auto; margin-right: auto;"'
)
```

---

*Data includes publications from `r min(student_data$Year)` to `r max(student_data$Year)`. Last updated: `r format(Sys.time(), "%B %d, %Y")`*